RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-29 manuscripts/manuscript.Rmd
Edits for Reading Phylogenetics Trees section.

Conceptually, a NeXML document has the following components: (1) phylogeny
data, (2) character or trait data in matrix form, (3) operational
taxonomic units (OTUs), and (4) metadata. When reading and writing NeXML,
`RNeXML` aims to map these to and from, respectively, their most
natural representations in R. These are the `phylo` object from the
`ape` package [@Paradis_2004], R's native `data.frame` list structure for
phylogenies, character matrices, and metadata, respectively.
<!-- compared to the preceding enumeration, this is missing where the OTUs go -->

@@

The current stable version of the `RNeXML` package can be installed from
the CRAN library[^1].

@@

@@

We begin by loading the `RNeXML` package and reading in an example
NeXML file provided with the package, containing two small phylogenetic
trees.[^2]

```{r}
library("RNeXML")
nex <- nexml_read("components/trees.xml")
```
The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
file. 
Objects of type `nexml` represent phylogenetic trees as well as 
character traits, and semantic metadata. As such, this new R object format differs
from existing formats such as `ape::phylo` format [@Paradis2004],
which describes trees alone, or any of the `phylobase` classes of object that contain 
trees and traits but more limited metadata [@phylobase].

Although new custom R functions and packages 
could be written for using `nexml` objects, for most users it will be
more convienent to use methods of the many already existing R packages for 
phylogenetics. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

```{r}
phy <- get_trees(nex)
```

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting function from
the `ape` package to plot the first tree in the resulting `ape::phylo` object returned by
`get_trees`.

@@

Note that NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function is designed for interactive use, and thus attempts to return the most sensible R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.

For scripted use, it is desirable to have a consistent return type.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/) 
 through the
`nexml` format.  `RNeXML`'s `nexml_read` can also read a file directly from a URL by automatically recognizing the URL patterns and first downloading the file to a working directory:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```

>>>>

Conceptually, a NeXML document has the following components: (1) phylogeny
data, (2) character or trait data in matrix form, (3) operational
taxonomic units (OTUs), and (4) metadata. When reading and writing NeXML,
`RNeXML` aims to map these to and from, respectively, their most
natural representations in R. These are the `phylo` object from the
`ape` package [@Paradis_2004], R's native `data.frame` list structure for
phylogenies, character matrices, and metadata, respectively.
<!-- compared to the preceding enumeration, this is missing where the OTUs go. HL -->

@@

As for other R packages, prior to using `RNeXML`, the package needs to
be installed and loaded. The current stable version can be installed from
the CRAN library[^1]:

@@

This will also install its dependencies. Once installed, R scripts
using the library only need to load it:

@@

We begin by reading in an example NeXML file provided with the
package, containing two small phylogenetic trees.[^2]

```{r}
nex <- nexml_read("components/trees.xml")
typeof(nex)
```

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml`. This object type is a representation of the NeXML
document contents, and therefore holds phylogenetic trees as well as 
character or trait matrices, and all metadata. This is similar in
design to the classes in the `phylobase` package, which contain 
trees and traits but hold only limited metadata [@phylobase], but
different from R object types such as `ape::phylo` (the `phylo` class
in the `ape` package [@Paradis2004]), which represents trees alone.

To interface conveniently with the many already existing R packages for 
phylogenetics, which conventionally accept phylogenies as `ape::phylo`
objects, the method `get_trees_list()` returns a `ape::multiPhylo`
object, which can be treated as a list of `ape::phylo` objects:

```{r}
phy <- get_trees_list(nex)
typeof(phy)
```

As an example for leveraging the wealth of resources in R
interoperating with the `ape::phylo` type, we use `ape`'s standard
plotting function to plot the first tree in the list of trees:

@@

The `get_trees_list()` method is designed for use in scripts, for
which a consistent and predictable return type (namely
`ape::multiPhylo`) is desirable.
<!-- the open (and previously not covered) issue here is what does -->
<!-- get_trees_list() return in the case of multiple trees blocks? HL -->
Following a common design practice
for R packages, `RNeXML` also contains a method `get_trees()`, which
is designed for interactive use and returns the R object most
intuitive given the arrangement of phylogeny data in the source NeXML
document. NeXML inherits from NEXUS the ability to contain multiple
phylogenies, and to group sets of phylogenies into "blocks". The
`get_trees()` method will return an `ape::phylo` object if the NeXML
document contains a single phylogeny, an `ape::multiPhylo` object if
it contains multiple phylogenies arranged in a single `trees` block,
and a list of `ape::multiPhylo` objects if it contains multiple
`trees` blocks. This design spares an interactive user from having to
type unnecessary data conversion commands, at the expense of needing
to know which return type to expect (or verifying the type via
introspection commands such as `typeof()`).

Because the example file supplied with the package contains two
phylogenies in a single `trees` block, in this case the return type of both
phylogeny accessor methods is the same:

```{r}
phy <- get_trees(nex)
typeof(phy)
plot(phy[[1]])
```

As mentioned above, the `nexml_read()` method allows specifying the
NeXML document to read not only as the name of a local file, but also
as the URL of a document on the web. If the location parameter is
recognized as a URL, the method will automatically download the
document to the local working directory and read it from there. This
gives convenient and rapid access to phylogenetic data published in
NeXML format on the web, which includes the content of the
phylogenetic data repository TreeBASE [@Piel_2002]. For
example, the following plots a tree in TreeBASE:

```{r eval=FALSE}
tb_nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
tb_phy <- get_trees_list(tb_nex)
plot(tb_phy[[1]])
```
